On the value of early warning systems for outbreak response

### 2023-06-13 paper/manuscript.qmd
Fix n in binomial declaration explaination to be a vector rather than a number
updating model, sample post-processing function

---
title: On the value of genomic wastewater surveillance information for pandemic response
subtitle: A model-based analysis
#author:
#  - name: Pedro Nascimento de Lima
#    email: plima@rand.org
#    affiliations: 
#        - id: rand
#          name: RAND Corporation
#          department: Engineering and Applied Sciences
#          address: 1776 Main Street
#          city: Santa Monica
#          state: CA
#          postal-code: 90401
#    attributes:
#        corresponding: true
#  - name: and colleagues
#    email: others@rand.org
#    affiliations:
#        - ref: rand
abstract: |
  While wastewater-based genomic surveillance (WBS) systems may provide an early warning system for pandemic response, its adoption is still incomplete and it is unclear whether public health departments should allocate a larger budget to those systems. This paper introduces a model to assess the value provided by wastewater genomic surveillance network and clarifies the conditions under which WBS systems provide the most value to public health decision makers.
keywords: 
  - Genomic Surveillance
  - Pandemic Response
# date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  elsevier-html:
    keep-tex: true
    mermaid:
      theme: neutral
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  html:
    keep-tex: true
    mermaid:
      theme: neutral
---

@@

where $B(\mathbf{n},p)$ results in a binomial draw from population vector $n$ with probability $p$. The model simulates outflows and inflows such that the outflows from one disease state are the same as the inflows to a different state.

@@

## Net monetary benefit (NMB)

The net monetary benefit of interventions can be expressed as

$$
E[NMB] = E[QALYs] * WTP - E[C_{npi}] - C_{surv} 
$${#eq-nbm}

where $QALYs$ represent the quality-adjusted life-years gained from non-pharmaceutical interventions, $WTP$ is the willingness-to-pay for one quality-adjusted life_year, $C_{npi}$ are the total costs of NPIs and $C_{surv}$ is the total cost of surveillance system employed.

<!--

# Valuing Wastewater surveillance information

In this analysis, WBS systems are not assumed to resolve uncertainty of an exogenous model parameter $\boldsymbol \theta$. If WBS systems did resolve those uncertainties, then it would be possible to estimate the value of information (VOI) provided by wastewater surveillance systems using an expected value of partial perfect information (EVPPI) approach. Rather, we assume that WBS systems change what signals are available to policymakers and hence *how* they respond to an endogenous model quantity. Hence, we define the value of wastewater surveillance information as simply the difference between the net monetary benefit of with and without surveillance

$$
V = E[NMB|WBS=1] - E[NMB|WBS=0]
$$ {#eq-wbsvalue}

-->

>>>>

---
title: On the value of early warning systems for outbreak response
subtitle: A model-based analysis
#author:
#  - name: Pedro Nascimento de Lima
#    email: plima@rand.org
#    affiliations: 
#        - id: rand
#          name: RAND Corporation
#          department: Engineering and Applied Sciences
#          address: 1776 Main Street
#          city: Santa Monica
#          state: CA
#          postal-code: 90401
#    attributes:
#        corresponding: true
#  - name: and colleagues
#    email: others@rand.org
#    affiliations:
#        - ref: rand
abstract: |
  While wastewater-based genomic surveillance (WBS) systems may provide an early warning system for pandemic response, its adoption is still incomplete and it is unclear whether public health departments should allocate a larger budget to those systems. This paper introduces a model to assess the value provided by wastewater genomic surveillance network and clarifies the conditions under which WBS systems provide the most value to public health decision makers.
keywords: 
  - Genomic Surveillance
  - Pandemic Response
# date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  elsevier-html:
    keep-tex: true
    mermaid:
      theme: neutral
    journal:
      #name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
  html:
    keep-tex: true
    mermaid:
      theme: neutral
  docx:
      keep-tex: true
---

@@

where $B(\mathbf{n},p)$ results in a binomial draw from population vector $\mathbf{n}$ with probability $p$. The model simulates outflows and inflows such that the outflows from one disease state are the same as the inflows to a different state.

@@

## Net monetary benefit (NMB) of the surveillance system

The total expected cost the outbreak for a given set of parameters $\mathbf{\theta}$ and surveillance system $i$ can be expressed as the sum of health costs $E[CH_{i}|\mathbf{\theta}]$, nonpharmaceutical intervention costs $E[CNPI_{ti}|\mathbf{\theta}]$ and surveillance costs $CSURV_{it}|\mathbf{\theta}$, where the expectation is taken across stochastic model realizations. Note that we chose to condition the expected cost of the pandemic on $\mathbf{\theta}$ rather than assuming a probability distribution over it.

$$
E[C_{i}|\mathbf{\theta}] = E[CH_{i}|\mathbf{\theta}] + \sum_t{E[CNPI_{ti}|\mathbf{\theta}] + CSURV_{it}|\mathbf{\theta} } 
$${#eq-costs}

Health costs $CH_{i}$ are computed as a function of number of removed individuals $R$ in the simulation
$$
E[CH_{i}|\mathbf{\theta}] = E[R(rw+o) |\mathbf{\theta}]
$${#eq-costs}
where $r$ is the infection fatality rate, $w$ is the average cost per death (assuming a set willingness to pay), and $o$ are other costs per infection, including days absent from work and any potential long-term disability.

In this analysis, WBS systems are not assumed not resolve uncertainty of any exogenous model parameter in $\boldsymbol \theta$. If WBS systems did resolve at least one of those uncertainties,  it would be possible to estimate the value of information (VOI) provided by wastewater surveillance systems using an expected value of partial perfect information (EVPPI) method. Rather, we assume that WBS systems change what signals are available to policymakers and hence *how* they respond to the pandemic. Hence, we define the net monetary benefit of wastewater surveillance information as the difference between the total cost of the pandemic with and without WBS surveillance.

$$
NMB_{surv} |\mathbf{\theta} = E[C_{conv surv}|\mathbf{\theta}]  - E[C_{meta surv}|\mathbf{\theta}]
$$ {#eq-wbsvalue}
